---
title: "Welcome to My Website"
---

## About Me

Hello, My name is Gisell Bennett! I'm a **Data Analytics major at Baruch College**, interested in exploring data and using it to drive meaningful insights. I'm currently learning various tools and techniques in data analysis, machine learning, and big data technologies.

## Skills I'm Learning

🚀 **Big Data & Cloud Computing**: AWS EMR, Apache Spark\

------------------------------------------------------------------------

*Thank you for visiting! Stay tuned for more updates on my data analytics journey.*

```{r}
#| include: false
1+1
```

------------------------------------------------------------------------

Last Updated: `r format(Sys.time(), "%A %m %d, %Y at %H:%M%p")`

```{r}
#| echo: false
#| message: false
#| warning: false

if(!require("leaflet")){
    options(repos=c(CRAN="https://cloud.r-project.org"))
    install.packages("leaflet")
    stopifnot(require("leaflet"))
}

baruch_longitude <- -73.98333
baruch_latitude <- +40.75028

leaflet() |>
  addTiles() |>
  setView(baruch_longitude, baruch_latitude, zoom=17) |>
  addPopups(baruch_longitude, baruch_latitude, 
            "I am a Master's student at <b>Baruch College</b>!")
```
